The Emergency Homeless Shelter Availability Across Toronto are Insufficient For Overwhelming Demand

### 2024-09-27 paper/paper.qmd
added graph sketches + edits, added entry id back to cleaned data

---
title: "Emergency Homeless Shelter Availabilities Across Toronto"
author: "Cher Ning-Li"
thanks: "Code and data are available at: <https://github.com/cher-ning/shelter_occupancies>"
date: 09-27-2024
date-format: long
abstract: "This paper analyzes the occupancy and capacity of emergency homeless shelters across Toronto, noting the high demand across all regions throughout the months. Dividing analysis of the data into different neighbourhoods, areas in Old Toronto are identified to have higher bed occupancies and number of shelters. Similarly, problem areas in North York and Old Toronto with patterns of inability to provide the number of bed spaces promised are also identified. Combining with the context of the current literature researching homelessness, these findings indicate a high urgency for the Toronto city council to implement effective short-term and long-term support programs to target the current homelessness crisis."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

\newpage
# Introduction

Much like many other large cities around the world, Toronto is facing a homelessness crisis that seems to be ever-increasing. Whether it is because of environmental factors such as rising housing prices and lack of employment opportunities, or other issues such as drug abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 [@cbcArticle]. The city council has recognized the urgency of this problem, declaring homelessness an emergency, and pledged their full efforts towards helping the homeless population move into safe homes [@cbcArticle]. Since then, the city of Toronto has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects. Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600-2,000 households access and transition into permanent housing in 2023-2024 [@cbcArticle].

In addition to these long-term plans that aim to move homeless individuals into stable homes, temporary shelters are also crucial during this transition process to provide short-term safety and resources. Research has shown that without access to shelter services, individuals can be pushed into substance-use relapses and hospitalizations, food insecurity, institutional circuitry through criminal justice systems, and other factors of instability [@citeKerman]. Shelters also provide useful data to track and understand the current state of the city's battle against this crisis.

With that in mind, how did the city council's efforts turn out? Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results. Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months [@shelterDataset]. Despite the evident high demand, there is even a decrease, albeit small, in shelters' funding capacity over the months [@shelterDataset].

In this paper, emergency shelter occupancy and availability will be compared across different months as well as different regions of Toronto to search for patterns or fluctutations in demand. The region of each shelter will be determined by its forward sorting area (FSA) code, or the first three digits of its postal code. As well, the regions that observe the highest frequency of failing to make efficient use of funding are identified as potential problem areas; further investigation into potential causes would be beneficial to understand if there are improvements that can be made so that the limited available funding can be maximized.

@@

The dataset used for analysis is of shelter occupancy in the year 2024 [@shelterDataset], from Open Data Toronto [@opendatatoronto]. The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License (@sec-license). However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records are accurate to their actual situation.

All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals [@shelterDataset]. With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them shelters compared to Transitional, this analysis will focus on Emergency programs only [@shelterDataset]. As well, all recorded shelters either measure capacity based on number of beds or rooms; similarly, this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset]. This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis. It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 was accessible as of writing this paper, and therefore only these entries were utilized in analysis.

@@

ggplot(occupancy_means, aes(x=month)) + 
  geom_line(aes(y = mean_occupied, colour = "Occupied Beds")) + 
  geom_line(aes(y = mean_capacity, colour="Funding Capacity")) +
  geom_line(aes(y = mean_actual, colour="Actual Capacity")) + 
  scale_colour_manual("", 
                      values = c("Occupied Beds"="darkred", "Funding Capacity"="steelblue", 
                                 "Actual Capacity"="darkblue")) +
  xlab(" ") +
  theme_classic() +
  labs(x = "Month", y = "Number of Beds") + 
  scale_x_continuous(breaks = seq(1, 9, by = 1))
```

\newpage
The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month. The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month. Notably, @fig-occupancy-vs-month also shows a significant decrease in funding around March, which will be further discussed in @sec-discussion. It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential. This is a potential area of concern, and its implications will also be discussed in @sec-discussion.

Another dimension to consider is the geographical region of the shelters. Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74. Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, North York, and around High Park [@citeFSA].
```{r}
#| label: fig-regionoccupancy
#| fig-cap: Average Bed Occupancy of Shelters in Each FSA Region
#| echo: false

@@

```{r}
#| echo: false
#| include: false
#| eval: true

ordered_region_occupancy <- region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ]
```

@@

```{r}
#| label: tbl-fsa-numshelters
#| eval: true
#| tbl-cap: "Number of Emergency Shelter Locations Reported in Each Toronto FSA"
#| echo: false

@@

fsa_shelters <- data %>% select(fsa, location_id) %>% group_by(fsa) %>% 
  summarise(numshelters = n_distinct(location_id))
fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
top_fsa_shelters <- fsa_shelters %>% top_n(3, numshelters)
knitr::kable(top_fsa_shelters, align ="c", col.names= c("FSA", "Number of Shelter Locations"))
```

@@

unavailable_beds <- unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]
#
top_unavailability <- unavailable_beds %>% top_n(10, mean_unavailable) %>% select(fsa, mean_unavailable, unavailable_rate)
knitr::kable(top_unavailability, 
             align ="c", 
             col.names= c("FSA", "Number of Unavailable Beds", 
                          "% of Funding Capacity Unavailable"))
```

@@

As well, consistent maintenance of these emergency services is also crucial in the meantime to support the current wellbeing of the homeless population. 
In order to improve these emergency programs which are already stretching government budgets to the maximum, problem areas such as M2N and M5G must be investigated and tackled.

## Limitations and Next Steps {#sec-discussion-limitations}
- not a view of the complete year bc its only september

@@

# Appendix {#sec-appendix}
## Data License Information {#sec-license}
"Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].
# Additional data details

>>>>

---
title: "The Emergency Homeless Shelter Availability Across Toronto are Insufficient For Overwhelming Demand"
author: "Cher Ning-Li"
thanks: "Code and data are available at: <https://github.com/cher-ning/shelter_occupancies>"
date: 09-27-2024
date-format: long
abstract: "This paper analyzes the occupancy and capacity of emergency homeless shelters across Toronto, noting the high demand across all regions throughout the months. Dividing analysis of the data into different neighbourhoods, areas in Old Toronto are identified to have higher bed occupancies and number of shelters. Similarly, problem areas in North York and Old Toronto with patterns of inability to provide the number of bed spaces promised are also identified. Combining with the context of the current literature researching homelessness, these findings indicate a high urgency for the Toronto city council to implement effective short-term and long-term support programs to target the current homelessness crisis."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

# Introduction {#sec-introduction}

Much like many other large cities around the world, Toronto is facing a homelessness crisis that seems to be ever-increasing. Whether it is because of societal factors such as rising housing prices and lack of employment opportunities, or personal issues such as substance abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 [@cbcArticle]. In response, the city council has declared homelessness an emergency and pledged their full efforts towards helping the homeless population move into safe homes [@cbcArticle]. Since then, the city has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects [@cbcArticle].

In addition to these long-term projects, temporary shelters are also crucial during this transition process to provide short-term safety and resources. Research has shown that without access to shelter services, individuals can be pushed into substance-use relapses and hospitalizations, food insecurity, and institutional circuitry through criminal justice systems [@citeKerman]. These shelters also provide useful data to track and understand the current progress of the city's battle against this crisis.

With that in mind, how did the city council's efforts turn out? Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results. Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months [@shelterDataset]. As well, the funding capacity of shelters was observed to decrease instead of increase, despite the high demand [@shelterDataset].

In this paper, emergency shelter occupancy and availability will be compared across different months as well as different regions of Toronto to search for patterns. The region of each shelter will be determined by its forward sorting area (FSA) code, or the first three digits of its postal code. As well, the regions that observe the highest frequency of failing to make efficient use of funding are identified as potential problem areas; further investigation into potential causes would be beneficial to understand if there are improvements that can be made so that the limited available funding can be maximized.

@@

The dataset used for analysis is of shelter occupancy in the year 2024 [@shelterDataset], from Open Data Toronto [@opendatatoronto]. The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License (@sec-license). However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records accurately reflect their state.

All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals [@shelterDataset]. With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them compared to Transitional type shelters, this analysis will focus on Emergency programs only [@shelterDataset]. As well, all recorded shelters either measure capacity based on number of beds or rooms, and this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset]. This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis. It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 was accessible as of writing this paper, and therefore only these entries were utilized in analysis.

@@

ggplot(occupancy_means, aes(x=month)) + 
  geom_line(aes(y = mean_occupied, colour = "Occupied Beds")) + 
  geom_line(aes(y = mean_capacity, colour="Funding Capacity")) +
  geom_line(aes(y = mean_actual, colour="Actual Capacity")) + 
  scale_colour_manual("", 
                      values = c("Occupied Beds"="darkred", 
                                 "Funding Capacity"="darkgreen", 
                                 "Actual Capacity"="steelblue")) +
  xlab(" ") +
  theme_classic() +
  labs(x = "Month", y = "Number of Beds") + 
  scale_x_continuous(breaks = seq(1, 9, by = 1))
```

\newpage
Shelter occupancy can also be viewed from the angle of mean funding capacity, actual capacity, and real number of occupied beds per month. The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month. Notably, @fig-occupancy-vs-month also shows a significant decrease in funding around March, which will be further discussed in @sec-discussion. It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential. This is a potential area of concern, and its implications will also be discussed in @sec-discussion.

Another dimension to consider is the geographical region of the shelters. Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread of values across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74. Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, around High Park, and North York respectively [@citeFSA].
```{r}
#| label: fig-regionoccupancy
#| fig-cap: Average Bed Occupancy of Shelters in Each FSA Region
#| echo: false

@@

```{r}
#| label: tbl-fsa-bedsoccupied
#| tbl-cap: Top 5 FSAs by Average Beds Occupied/Night
#| echo: false

ordered_region_occupancy <- 
  region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ] %>% 
  top_n(5, mean_occupied)

knitr::kable(ordered_region_occupancy, 
             align ="c", 
             col.names= c("FSA", "Mean Number of Beds Occupied"))
```

@@

```{r}
#| label: tbl-fsa-numshelters
#| eval: true
#| tbl-cap: "Top 6 FSAs by Number of Emergency Shelter Locations"
#| echo: false

@@

fsa_shelters <- data %>% select(fsa, location_id) %>% group_by(fsa) %>% 
  summarise(numshelters = n_distinct(location_id))
fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
top_fsa_shelters <- fsa_shelters %>% top_n(3, numshelters)

knitr::kable(top_fsa_shelters, 
             align ="c", 
             col.names= c("FSA", "Number of Shelter Locations"))
```

@@

unavailable_beds <- 
  unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]

top_unavailability <- unavailable_beds %>% top_n(10, mean_unavailable) %>% 
  select(fsa, mean_unavailable, unavailable_rate)

knitr::kable(top_unavailability, 
             align ="c", 
             col.names= c("FSA", "Number of Unavailable Beds", 
                          "% of Funding Capacity Unavailable"))
```

@@

In the meantime, consistent maintenance of these emergency services is also crucial to support the current wellbeing of the homeless population. Temporary shelters, though unable to solve the core of the problem, are vital to preventing substance abuse relapses, suicidality, and further reduction of a support network and community connections [@citeKerman]. In order to improve these emergency programs which are already stretching government budgets to the maximum, problem areas such as M2N and M5G must be investigated and tackled.

## Limitations and Next Steps {#sec-discussion-limitations}
One major limitation of the current dataset is that it only covers data from January to September 2024 as of the writing of this paper. This means that any patterns present in the winter months of November and December are not observable to the current analysis. As well, due to the dataset being unaudited, its accuracy to the true situations at these shelter locations cannot be verified.

Further research should consider focussing in on the areas of higher occupancy and higher unavailabilities to understand the factors that may be causing these patterns, allowing for policy makers to better address the exact needs of the homeless population.

@@

# Appendix {#sec-appendix}

## Dataset and Graph Sketches {#sec-sketches}
Sketches created when planning out the desired dataset and graphs for the current paper are included in the Github repository.

## Data License Information {#sec-license}
"Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].

## Data Cleaning {#sec-datacleaning}
@tbl-sampledata displays 5 rows of the cleaned data used for analysis in this paper. The process involved filtering for only columns with variables of interest, removing rows with negative or NA entries, and renaming the columns for clarity and concision. Additional columns for month and FSA were also created for ease of grouping during analysis. Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were used.

```{r}
#| label: tbl-sampledata
#| eval: true
#| tbl-cap: "First 5 Rows of Cleaned Data"
#| echo: false

sample_data <- data %>% top_n(-5, id) %>% 
  select(month, date, fsa, location_id, actual_capacity, funding_capacity, 
         occupied_beds, unoccupied_beds, unavailable_beds)
knitr::kable(sample_data, 
             align = "c",
             col.names = c("Month", "Date", "FSA", "Location ID", 
                           "Actual Capacity", "Funding Capacity", 
                           "Occupied Beds", "Unoccupied Beds", 
                           "Unavailable Beds"))
```
